\thetitle

### 2009-02-19 paper/paper.tex
paper: Fix inverted inequality.
Ignore-this: d6a98221868499b0a1f6ff1a00f269fa

darcs-hash:20090219084636-bab43-62012aace058b21a63e5711102bae7872c12bde1.gz
james-final-tweaks
darcs-hash:20090219132958-d98fc-d81cd9874759ac4b5fafca435eef76905f7f3dad.gz

The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning. The representation we use takes the form of a shallow monadic embedding, in which the algorithm is written as a monadically expressed functional program. This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.

@@

Since we are working in Coq, we already have at our disposal a full functional
programming language, in the form of Coq's pCIC \cite{coq'art}. However, just
writing the algorithm as an ordinary Coq function would not let us observe its
use of comparisons. 
We can however see comparison counting as a \emph{side effect}. As is well
known and standard practice in functional languages such as Haskell, side
effects can be represented using \emph{monads}: a side-effecting function |f|
from |A| to |B| is represented as a function |A -> M B| where |M| is a type
constructor encapsulating the side effects. ``Identity'' and ``composition''
for such functions are given by |ret| (for ``return'') of type |A -> M A| and
|bind| (infix: |>>=|) of type |M A -> (A -> M B) -> M B| satisfying certain
identities (the \emph{monad laws}). For a general introduction to monadic
programming and monad laws, see \cite{wadler93monads}. Furthermore, we use
Haskell's ``do-notation'',
\begin{code}
 Notation "x <- y ; z" := (bind y (fun x: _ => z)) 
\end{code}
and freely use standard monadic functions such as:
\begin{code}
 liftM: forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)
 filterM: forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)
\end{code}
|Monad| is a dependent record containing the (coercible) carrier of type |Set
-> Set|, along with the |bind| and |ret| operations, and proofs of the three monad laws. 
% We make those uses observable by writing the algorithm as a monadically
% expressed function, parameterizing it on both the monad itself and on the
% comparison operation.

@@

We use Coq's |Program Fixpoint| facility \cite{subsetcoercions} to cope with Quicksort's non-structural recursion, specifying list length as an input measure function that is separately shown to strongly decrease for each recursive call. For |qs|, these proof obligations are trivial enough for Coq to prove mostly by itself.

For recursive functions defined this way, Coq does not automatically define corresponding induction principles matching the recursive call structure. Hence, for this |qs| definition as well as the one we will introduce in section \ref{nondetexpec}, we have defined these induction principles manually. To make their use as convenient as possible, we further customized and specialized them to take advantage of specific monad properties. We will ignore these issues in this paper, and will henceforth simply say: ``by induction on |qs|, ...''.

@@

We now expand the technique in preparation for the average-case complexity proof.

@@

These two modifications together greatly simplify the formalisation, because
they remove the need to carefully track input distributions in order to show
that nice inputs (for which the original deterministic version of the
algorithm performs well) sufficiently outnumber bad inputs (for which the
original version performs poorly). They further ensure that the $O(n \log n)$
average-case bound holds not just averaged over all possible input lists, but
for each individual input list as well. In particular, it means that once we
prove that the bound holds for an arbitrary input, the global bound
immediately follows.

@@

Note that the deterministic Quicksort definition in section \ref{embed} could have been implemented with a |partition| pass as well, which might well have made the worst-case proof even simpler. We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.

Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation. A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and a deterministic |pick| operation, like |head|.

Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program like
\begin{code}
  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2].
\end{code}
When executed in the list monad (commonly used to emulate nondeterministic computation), this program
produces |[0, 1, 2]| as its list of possible outcomes. Unfortunately, the information that 0 is a more likely outcome than 1 or 2, has been lost. Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|. This makes list nondeterminism unsuitable for our purposes.

@@

The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof could be fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.

@@

\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
    (forall x `In` q, expec (count x) t <= fr x) ->
    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
\end{code}
We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using identity \ref{expec_bind_lem} from section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
\begin{code}
  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0,
\end{code}
which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.

The second subgoal, expressed with |monoid_expec|, reads
\begin{code}
  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
\end{code}
This corresponds exactly to the property described at the beginning of this section. We prove it in the next section.

@@

Again, the proof is by induction on |qs|. But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
\begin{code}
  forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
    monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
\end{code}
In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
\begin{code}
monoid_expec (count (i, j)) (
    pi <- pick [0 ... n-1];
    let pivot := nth li pi in
    part <- partition_U pivot (remove li pi);
    lower <- qs_U (part Lt);
    upper <- qs_U (part Gt);
    ret (lower ++ pivot :: part Eq ++ upper)
  ) <= ecc i j.
\end{code}
Since |cmp_U| is deterministic, |partition_U| is as well. Furthermore, since
we know exactly what monadic effects |partition_U| has, we can split those
effects off and revert to simple effect-free |filter| passes. Finally, we rewrite using the following |monoid_expec| identity:
\begin{code}
  monoid_expec f (pick l >>= m) = avg (map (monoid_expec f fcmp m) l).
\end{code}
This way, the goal ends up in a form using less monadic indirection:
\begin{code}
avg (map (monoid_expec (count (i, j)) fcmp (fun pi =>
    let pivot := nth li pi in
    let rest := remove li pi in
    let flt := fun c => filter ((= c) fcmp lookup_cmp pivot) rest in
    ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
      lower <- qs_U (flt Lt);
      upper <- qs_U (flt Gt);
      ret (lower ++ pivot :: flt Eq ++ upper)
    ))) [0...n-1]) <= ecc i j.
\end{code}
Here, |map_fst| applies a function to a pair's first component.

@@

\begin{code}
monoid_expec (count (i, j)) (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
  <= ecc i j.
\end{code}
All indices in the first filtered list denote elements that are less than the element denoted by the pivot. Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|. And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property shown at the end of the previous section). Hence, the first |monoid_expec| term vanishes as well, leaving
\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= ecc i j.
\end{code}
We now compare |nth (sort l) i| with |nth (sort l) pivot|.
\begin{itemize}
\item If the two are equal, then |i| will not occur in the |filter| term, and so (again) no |(i, j)| comparisons are performed.
\item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
\item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis. For this, it must be shown that filtering the list of indices preserves contiguity, which follows from the fact that the indices share the order of the elements they denote in |sort l|.
\end{itemize}
This concludes the case where |i < pivot|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.

@@

We have shown that it is straightforward to give a fully formal treatment in type theory of a classical result in complexity theory. This clearly shows the utility and applicability of the general monadic framework for worst- and average-case complexity that we have developed.

>>>>

The key to giving a proper formal treatment of both of these aspects lies in the use of an appropriate representation of the algorithm, capable of capturing its computational behavior---specifically, its use of comparisons---in a way suitable for subsequent formal reasoning. 
The appproach we take is to consider such operation-counting as a \emph{side effect}, and to use the general framework of \emph{monads} for representing side-effecting computation in pure functional languages. 
The representation we use takes the form of a shallow monadic embedding, in which the algorithm, here Quicksort, is written as a monadically expressed functional program. This definition is then instantiated with refinements of operation-counting monads to make the comparison count observable.

@@

Since we are working in Coq, we already have at our disposal a full functional
programming language, in the form of Coq's pCIC \cite{coq'art}. However, just
writing the algorithm as an ordinary Coq function would not let us observe its
use of comparisons. 
We can however see comparison counting as a \emph{side effect}. As is well
known and standard practice in functional languages such as Haskell, side
effects can be represented using \emph{monads}: a side-effecting function |f|
from |A| to |B| is represented as a function |A -> M B| where |M| is a type
constructor encapsulating the side effects. ``Identity'' and ``composition''
for such functions are given by |ret| (for ``return'') of type |A -> M A| and
|bind| (infix: |>>=|) of type |M A -> (A -> M B) -> M B| satisfying certain
identities (the \emph{monad laws}). For a general introduction to monadic
programming and monad laws, see \cite{wadler93monads}. Furthermore, we use
Haskell's ``do-notation'',
\begin{code}
 Notation "x <- y ; z" := (bind y (fun x: _ => z)) 
\end{code}
and freely use standard monadic functions such as:
\begin{code}
 liftM: forall (M: Monad) (A B: Set), (A -> B) -> (M A -> M B)
 filterM: forall (M: Monad) (A: Set), (A -> M bool) -> list A -> M (list A)
\end{code}
Here, the Coq type |Monad| is a dependent record containing the (coercible) carrier of type |Set
-> Set|, along with the |bind| and |ret| operations, and proofs of the three monad laws. 
% We make those uses observable by writing the algorithm as a monadically
% expressed function, parameterizing it on both the monad itself and on the
% comparison operation.

@@

We use Coq's |Program Fixpoint| facility \cite{subsetcoercions} to cope with Quicksort's non-structural recursion, specifying list length as an input measure function that is separately shown to strongly decrease for each recursive call. For this definition of |qs|, these proof obligations are trivial enough for Coq to prove mostly by itself.

For recursive functions defined this way, Coq does not automatically define corresponding induction principles matching the recursive call structure. Hence, for this |qs| definition as well as the one we will introduce in section \ref{nondetexpec}, we had to define these induction principles manually. To make their use as convenient as possible, we further customized and specialized them to take advantage of specific monad properties. We will omit further discussion of these issues in this paper, and will henceforth simply say: ``by induction on |qs|, ...''.

@@

We now extend the technique to prepare for the average-case complexity proof.

@@

These two modifications together greatly simplify the formalisation, because
they remove the need to carefully track input distributions in order to show
that `good' inputs (for which the original deterministic version of the
algorithm performs well) sufficiently outnumber `bad' inputs (for which the
original version performs poorly). They further ensure that the $O(n \log n)$
average-case bound holds not just averaged over all possible input lists, but
for each individual input list as well. In particular, it means that once we
prove that the bound holds for an arbitrary input, the global bound
immediately follows.

@@

Note that the deterministic Quicksort definition in section \ref{embed} could also have been implemented with a |partition| pass instead, which might well have made the worst-case proof even simpler. We chose not to do this, in order to emphasize that the properties the average-case proof demands of the algorithm rule out the na{\"\i}ve but familiar implementation using |filter| passes.

Nondeterminism can now be emulated by instantiating these definitions with a suitable monad and |pick| operation. A deterministic, non-instrumented version can still be obtained, simply by using the identity monad and any deterministic |pick| operation, such as |head| or 'median-of-three'.

Let us now consider what kind of nondeterminism monad would be suitable for reasoning about the expected value of a nondeterministic program like
\begin{code}
  x <- pick [0, 1]; if x = 0 then ret 0 else pick [1, 2].
\end{code}
When executed in the list monad (commonly used to emulate nondeterministic computation), this program
produces |[0, 1, 2]| as its list of possible outcomes. Unfortunately, the information that 0 is a more likely outcome than 1 or 2 has been lost. Such relative probabilities are critical to the notion of an expected value: the expected value of the program above is |avg [0, avg [1, 2]] = {-"\frac{3}{4}"-} /= 1 = avg [0, 1, 2]|. This makes list nondeterminism unsuitable for our purposes.

@@

The right inequality is a strictly numerical affair, requiring a bit of analysis involving the harmonic series. As stated before, this part of the proof 
    was 
%%% could be 
fairly directly transcribed from the paper proof, with few complications and additions, and so we will not discuss it.

@@

\begin{code}
  forall (X: Set) (fr: X -> R) (q: list X) (t: ne_tree (list X)),
    (forall x `In` q, expec (count x) t <= fr x) ->
    (forall x `NotIn` q, expec (count x) t = 0) -> expec length t <= {-"\displaystyle{\sum_{\Varid{x} \in \Varid{q}}\Varid{fr}\ \Varid{x}}"-}.
\end{code}
We end up with two subgoals, the first of which is \begin{code}forall (i, j) `NotIn` IJ, expec (count (i, j)) (ne_tree.map fst (qs_U li)) = 0.\end{code} Rewriting this using identity \ref{expec_bind_lem} from section \ref{nondetexpec} in reverse, then rewriting the |expec| as a |monoid_expec|, and then generalizing the premise, results in
%%% this needs a label!!! (\label{monoid_expec_zero_lem})
\begin{code}
  forall i j li, (i `NotIn` li \/ j `NotIn` li) -> monoid_expec (count (i, j)) (U.qs li) = 0 
\end{code}
which can be shown by induction on |qs|, although we will not do so in this paper. We \emph{will} use this property again in the next section.

The second subgoal, expressed with |monoid_expec|, becomes
%%% this needs a label!!! (\label{monoid_expec_IJ_lem})
\begin{code}
  forall (i, j) `In` IJ, monoid_expec (count (i, j)) (qs_U li) <= ecc i j. 
\end{code}
which corresponds exactly to the property described at the beginning of this section. We prove it in the next section.

@@

Again, the proof %%% of (\ref{monoid_expec_IJ_lem}) 
is by induction on |qs|. But to get a better induction hypothesis, we drop the |(i, j) `In` IJ| premise (because as was shown in the last section, the statement is also true if |(i, j) `NotIn` IJ|), and add a premise saying |li| is a permutation of a contiguous sequence of indices.
\begin{code}
  forall i j, i < j -> forall (li: list nat) (b: nat), Permutation [b ... b + length li - 1] li ->
    monoid_expec (count (i, j)) (qs_U li) <= ecc i j.
\end{code}
In the base case, |li| is |nil|, and the left side of the inequality reduces to 0. In the recursive case, |qs| unfolds:
\begin{code}
monoid_expec (count (i, j)) (
    pi <- pick [0 ... n-1];
    let pivot := nth li pi in
    part <- partition_U pivot (remove li pi);
    lower <- qs_U (part Lt);
    upper <- qs_U (part Gt);
    ret (lower ++ pivot :: part Eq ++ upper)
  ) <= ecc i j.
\end{code}
Since |cmp_U| is deterministic, |partition_U| is as well. Furthermore, since
we know exactly what monadic effects |partition_U| has, we can split those
effects off and revert to simple effect-free |filter| passes. Finally, we rewrite using the following |monoid_expec| identity:
\begin{code}
  monoid_expec f (pick l >>= m) = avg (map (monoid_expec f fcmp m) l).
\end{code}
This way, the goal ends up in a form using less monadic indirection:
\begin{code}
avg (map (monoid_expec (count (i, j)) fcmp (fun pi =>
    let pivot := nth li pi in
    let rest := remove li pi in
    let flt := fun c => filter ((= c) fcmp lookup_cmp pivot) rest in
    ne_tree.map (map_fst (++ map (unordered_nat_pair pivot) rest)) (
      lower <- qs_U (flt Lt);
      upper <- qs_U (flt Gt);
      ret (lower ++ pivot :: flt Eq ++ upper)
    ))) [0...n-1]) <= ecc i j.
\end{code}
Here, |map_fst| applies a function to a pair's first component.

@@

\begin{code}
monoid_expec (count (i, j)) (qs_U (filter ((= Lt) fcmp lookup_cmp pivot) rest)) +
monoid_expec (count (i, j)) (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest))
  <= ecc i j.
\end{code}
All indices in the first filtered list denote elements that are less than the element denoted by the pivot. Since the former precede the latter in |sort l|, it must be the case that these indices are all less than |pivot|. And since |pivot < i|, it follows that the first |qs_U| term will produce no |(i, j)| comparisons (using the property 
%%%(\ref{monoid_expec_IJ_lem}) 
shown at the end of the previous section). Hence, the first |monoid_expec| term vanishes as well, leaving
\begin{code}
monoid_expec (count (i, j))
  (qs_U (filter ((= Gt) fcmp lookup_cmp pivot) rest)) <= ecc i j.
\end{code}
We now compare |nth (sort l) i| with |nth (sort l) pivot|.
\begin{itemize}
\item If the two are equal, then |i| will not occur in the |filter| term, and so (again) no |(i, j)| comparisons are performed.
\item If |nth (sort l) i < nth (sort l) pivot|, then we must have |i < pivot|, contradicting the assumption that |pivot < i|.
\item If |nth (sort l) i > nth (sort l) pivot|, then we apply the induction hypothesis. For this, it must be shown that filtering the list of indices preserves contiguity, which follows from the fact that the indices share the order of the elements they denote in |sort l|.
\end{itemize}
This concludes the case where |pivot < i|. The case where |j < pivot| is symmetric. The other three cases use similar arguments.

@@

We have shown that it is straightforward to give a fully formal treatment in type theory of a classical result in complexity theory. This clearly shows the utility and applicability of the general monadic 
%%%framework for 
approach to worst- and average-case complexity that we have developed.
